---
title: 'TIP-RA Baseline: ACPA Presence Analysis For Paper'
author: "Lauren Vanderlinden"
date: "Report generated: `r paste(format(Sys.Date(), '%B %d, %Y'))`"
output:
  html_document:
    number_sections: no
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float: yes
---

# Introduction

This is updating the unstratified analysis performed for the paper.  We previously used a continuous measure for ACPA as the outcome, and here we will be modeling ACPA as either positive or negative.  

## ACPA Positivity 

Instead of modeling ACPA abundance, we are going to model ACPA presence (yes or no).  The criteria to determine ACPA positivity will be defined as the mean + 3 SD based on the control (CCP negative) group.  Apply this across all subjects to get a positive/negative call for each ACPA peptide.  **Note:**  Not going to perform this on the aggregate measures.    

## Analysis Approach 

Will perform a **logistic regression** using group (either CCP-, CCP+ or early RA) to predict ACPA positivity adjusting for age, sex and ever a smoker status.  Multiple testing adjustment was accounted for using the Bemjamini Hochberg false discovery rate (FDR) method on the F-test p-value for group effect.  Pairwise comparisons are reported.  


```{r, include=FALSE}
rm(list=ls())

library(readxl)
library(kableExtra)
library(ggplot2)
library(RColorBrewer)
#library(ggbiplot)
library(factoextra)
#library(WGCNA)
library(emmeans)
library(table1)
library(DT)
library(multcomp)
library(dplyr)
library(ggnewscale)
library(circlize)
library(ComplexHeatmap)

```

# Demographics

This is the demographic table used for the analysis.  This is just for reference and the total N's for these groups are what I added to supplemental table 5.  For the N's in supplemental table 6, I looked at the N's within the '0401Count'.      

```{r, warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
wd = "R:/TIP-RA/LaurenV/baselineCrossSection/"
load(file=paste0(wd, "data/forBaselineAnalysis.Rdata"))

#in array matrix, remove the final 2 columns (CCP3Final and Converter) & also put in the aggregate (sum)
array2= array[,-c(43:44)]
sum.all = apply(array2, 1, sum)
sum.all.log = log(sum.all)

cilpArray = array2[,grep("CILP", colnames(array2))]
sum.citCILP = apply(cilpArray[,grep("cit", colnames(cilpArray))], 1, sum)
array3 = cbind(array2, sum.citCILP, sum.all, sum.all.log) 

pheno.want = pheno[which(pheno$LabId %in% array$Labid),]
pheno.want = pheno[,which(colnames(pheno) %in% c("LabId", "Analyze", "SE", "SECount", "0404Count", "0401Count", "Gender", "Age", "EverSmoke", "CurrentSmoker"))]

forAnalysis = merge(pheno.want, array3, by.y="Labid", by.x="LabId")
forAnalysis = forAnalysis[-which(forAnalysis$Analyze %in% c("BaselineReverter", "Do Not Analyze")),]

forAnalysis$sum0401_0404 = forAnalysis$`0401Count`+ forAnalysis$`0404Count`
forAnalysis$either0401_0404 = as.numeric(forAnalysis$sum0401_0404 > 0)

###Look at the distribution of genotype variables;
table1(~ SE + as.factor(SECount) + as.factor(`0401Count`) + EverSmoke + Gender + Age | Analyze, dat=forAnalysis)
```

# Unstratified ACPA Positivity Analysis

This is the basic logistic model with group (either CCP-, CCP+ or early RA) predicting ACPA positivity while adjusting for age, sex and ever smoker status.

## Criteria Distribution

Just a quick glance at how distriution of different positivity crtieria.  So depending on the ACPA peptide you are looking at, we could have a positivity criteria ranging from 496 to 4,092.  

```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.height = 2.5, fig.width = 3.5}
ACPAvars = as.matrix(colnames(forAnalysis)[c(11:51)])
ACPA = ACPAvars[1,1]
### Get ACPA positivity criteria:
getPosCriteria = function(ACPA){
  tmp = forAnalysis[which(forAnalysis$Analyze=="CCP_Neg_Baseline"),]
  colnames(tmp)[which(colnames(tmp)==ACPA)] = "outcome"
  mean_outcome = mean(tmp$outcome)
  sd_outcome = sd(tmp$outcome)
  criteria_3SD = mean_outcome + 3*sd_outcome
  criteria_2.5SD = mean_outcome + 2.5*sd_outcome
  criteria_2SD = mean_outcome + 2*sd_outcome
  
  want = data.frame(ACPA, criteria_3SD, criteria_2.5SD, criteria_2SD)
  return(want)
}

criteria = do.call(rbind, apply(ACPAvars, 1, function(a) getPosCriteria(a)))

ggplot(criteria, aes(x=criteria_3SD)) + 
  geom_histogram(color="black", fill="steelblue")+
  theme_bw()+
  xlab("ACPA Positivity Criteria")
```

## Results Table 

I'm showing the number of ACPA positive (%) in each group and the p-values.  This is similar to the means and p-values reported in other supplemental tables. 

If you want a table with odds ratios, 95% CI as well, let me know and I'll create that.  
```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.height = 2.5, fig.width = 3.5}

ACPA_cont = forAnalysis[,which(colnames(forAnalysis) %in% ACPAvars)] 
rownames(ACPA_cont) = forAnalysis$LabId

getPos = function(ACPA, criteria_want){
  cut = criteria[which(criteria$ACPA==ACPA), paste0("criteria_", criteria_want)]
  ACPA_pos = as.numeric(ACPA_cont[,which(colnames(ACPA_cont)==ACPA)]>=cut)
  return(ACPA_pos)
}

ACPA_pos_3SD = apply(ACPAvars, 1, function(a) getPos(a, "3SD"))
rownames(ACPA_pos_3SD) = rownames(ACPA_cont)
colnames(ACPA_pos_3SD) = ACPAvars

covsWant = forAnalysis[,which(colnames(forAnalysis) %in% c("LabId", "Analyze", "Gender", "Age", "SE", "EverSmoke"))]

forAnalysis_3SD = merge(covsWant, ACPA_pos_3SD, by.x="LabId", by.y=0)

getResults_3SD = function(ACPA){
  #get new datasets
    tmp = forAnalysis_3SD
    colnames(tmp)[which(colnames(tmp)==ACPA)]="out"
    tmp$Analyze = as.factor(tmp$Analyze)

    #model
    
    mod = glm(out ~ Analyze + Gender + Age + EverSmoke, data = tmp, family="binomial")
    mod_null = glm(out ~ Gender + Age + EverSmoke, data = tmp, family="binomial")
    mod_pairwise = glht(mod, mcp(Analyze="Tukey"))
    overall = anova(mod, mod_null, test="LRT")

    #extract different p-values
    pval_overall = overall[2,5]
    pval_neg_vAtRisk = summary(mod_pairwise)$test$pvalues[1]
    pval_atRisk_vRA = summary(mod_pairwise)$test$pvalues[3]
    pval_neg_vRA = summary(mod_pairwise)$test$pvalues[2]

    toReport = c(pval_neg_vAtRisk, pval_atRisk_vRA, pval_neg_vRA, pval_overall) 
    names(toReport) = c("CCP+ vs CCP-", "RA vs CCP+", "RA vs CCP-", "overall p-value")
    return(toReport)
}

results_pvals = t(apply(ACPAvars, 1, function(a) getResults_3SD(a)))
rownames(results_pvals) = ACPAvars[,1]

results_pvals = as.data.frame(results_pvals)
results_pvals$FDR_CCPpos_vNeg = p.adjust(results_pvals$`CCP+ vs CCP-`, method="BH")

getNs = function(ACPA){
  tmp = forAnalysis_3SD
  colnames(tmp)[which(colnames(tmp)==ACPA)]="out"
  Ns = table(tmp$out, tmp$Analyze)
  N_1 = Ns[2,]
  percents = paste0("(", round(100*Ns/colSums(Ns), 2)[2,], "%)")
  Ns_pos_per = cbind(N_1, percents)
  want = apply(Ns_pos_per, 1, function(a) paste0(a[1], " ", a[2]))
  return(want)
}

toPrint = t(apply(ACPAvars, 1, function(a) getNs(a)))
rownames(toPrint) = ACPAvars

results_pvals$FDR_RA_vCCPpos = p.adjust(results_pvals$`RA vs CCP+`, method="BH")

results_pvals$RA_vCCP_pretty = paste0(formatC(results_pvals$`RA vs CCP+`, format="e", digits=2), " (", round(results_pvals$FDR_RA_vCCPpos, 3), ")")
results_pvals$CCP_vNeg_pretty = paste0(formatC(results_pvals$`CCP+ vs CCP-`, format="e", digits=2), " (", round(results_pvals$FDR_CCPpos_vNeg, 3), ")")
results_pvals$overall_pretty = formatC(results_pvals$`overall p-value`, format="e", digits=2)
  
toPrint2 = cbind(toPrint, results_pvals[,c(9,8,7)])
colnames(toPrint2) = c("Anti-CCP3", "At-Risk", "Early RA", "overall p-value", "At-Risk vs Anti-CCP3 p-value (FDR)", "Early RA vs At-Risk p-value (FDR)")
datatable(toPrint2,rownames=TRUE)
```

## Candidate 

There is a single ACPA peptide that passes a FDR adjusted p-value < 0.05: **H4 33-48 cit39 40 (79)**

Lets look at the bar charts for this peptide:

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 5}

toPlot = as.data.frame(table(forAnalysis_3SD$Analyze, forAnalysis_3SD$`H4 33-48 cit39 40 (79)`))
colnames(toPlot) = c("Group", "H4 33-48 cit39 40 (79) Positivity", "N Subjects")
toPlot$Group = gsub("CCP_Neg_Baseline", "Anti-CCP3", toPlot$Group)
toPlot$Group = gsub("CCP_Pos_Baseline", "At-Risk", toPlot$Group)
toPlot$Group = gsub("RA_Baseline", "Early RA", toPlot$Group)

ggplot(data=toPlot, aes(x=`Group`, y=`N Subjects`, fill=`H4 33-48 cit39 40 (79) Positivity`)) +
geom_bar(stat="identity", position=position_dodge())+
  theme_bw()+
  geom_text(aes(label=`N Subjects`), position=position_dodge(width=0.9), vjust=-0.25, size=3)+
  scale_fill_manual(values=c("gray", "darkred"))+
  labs(title="H4 33-48 cit39 40 (79) Positivity") +
  ylab("Number of Subjects")+
  theme(legend.position="bottom")
```

## Heatmap for Paper

Probably just use Kevin's, but here it one I made for the report.  

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 5, eval=FALSE}

head(forAnalysis_3SD)
write.csv(forAnalysis_3SD, file="R:/TIP-RA/LaurenV/baselineCrossSection/data/forPaper/ACPA_meanPlus3SDyesNo.csv", row.names=FALSE)
```
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}
toPlot = forAnalysis_3SD[order(forAnalysis$Analyze),]
rownames(toPlot) = toPlot$LabId
toPlot = toPlot[,-c(1, 3:6)]

toPlot.neg = as.matrix(t(toPlot[which(toPlot$Analyze=="CCP_Neg_Baseline"),-1]))
toPlot.pos =  as.matrix(t(toPlot[which(toPlot$Analyze=="CCP_Pos_Baseline"),-1]))
toPlot.RA =  as.matrix(t(toPlot[which(toPlot$Analyze=="RA_Baseline"),-1]))

hp = Heatmap(toPlot.neg, col=c("darkblue", "red"), column_title="Anti-CCP3\n(n=172)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10)) + 
Heatmap(toPlot.pos, col=c("darkblue", "red"), column_title="At-Risk\n(n=97)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10)) +
Heatmap(toPlot.RA, col=c("darkblue", "red"), column_title="Early RA\n(n=62)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10))
hp
#png(file="R:/TIP-RA/LaurenV/baselineCrossSection/manuscript/ACPA_heatmap.png", res=600,width=480*12, height=480*7)
#hp
#dev.off()
```

# Stratified ACPA Positivity Analysis

Here we are going to stratify the dataIn by **0401 positivity**.  In the dataset, we have it counted as either 0, 1, or 2.  But we are stratifying into 2 groups: either 0 or at least 1.    

## Distribution of 0401 genotype

Here is the table of the 0401 genotype in our dataset.  It is classified as either 0, 1 or 2 copies of the 0401 allele.  But as we can see the 2 copies are fairly rare, so we are going to stratify the data by 0 copies or at least 1 copy of the 0401 allele.   

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}
forAnalysis$geno0401_bi = as.numeric(forAnalysis$`0401Count`!=0)
toPrint = table(forAnalysis$`0401Count`, forAnalysis$geno0401_bi)
toPrint = as.data.frame(cbind(rownames(toPrint), toPrint))
colnames(toPrint) = c("Number of 0401 Alleles", "0401 Negative", "0401 Positive")
kable(toPrint, row.names=F, align='l') %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) 
```
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}
geno= "geno0401_bi"
pheno.geno = forAnalysis[,which(colnames(forAnalysis) %in% c("LabId", "geno0401_bi", "0401Count"))]

forAnalysis_3SD = merge(pheno.geno, forAnalysis_3SD, by="LabId")

getStratifiedResults = function(ACPA){
    tmp = forAnalysis_3SD
    colnames(tmp)[which(colnames(tmp)==ACPA)]="out"
    tmp$Analyze = as.factor(tmp$Analyze)
    tmp.neg = tmp[which(tmp$geno0401_bi==0),]
    tmp.pos = tmp[which(tmp$geno0401_bi==1),]
  
    ###positive model ###
    mod = glm(out ~ Analyze + Gender + Age + EverSmoke, data = tmp.pos, family="binomial")
    mod_null = glm(out ~ Gender + Age + EverSmoke, data = tmp.pos, family="binomial")
    mod_pairwise = glht(mod, mcp(Analyze="Tukey"))
    overall = anova(mod, mod_null, test="LRT")

    #extract different p-values
    pval_overall = overall[2,5]
    pval_neg_vAtRisk = summary(mod_pairwise)$test$pvalues[1]
    pval_atRisk_vRA = summary(mod_pairwise)$test$pvalues[3]
    pval_neg_vRA = summary(mod_pairwise)$test$pvalues[2]

    toReport = c(pval_neg_vAtRisk, pval_atRisk_vRA, pval_neg_vRA, pval_overall) 
    names(toReport) = c("Positive: CCP+ vs CCP-", "Positive: RA vs CCP+", "Positive: RA vs CCP-", "Positive: overall p-value")
    
    ###negative model ###
    mod.n = glm(out ~ Analyze + Gender + Age + EverSmoke, data = tmp.neg, family="binomial")
    mod_null.n = glm(out ~ Gender + Age + EverSmoke, data = tmp.neg, family="binomial")
    mod_pairwise.n = glht(mod.n, mcp(Analyze="Tukey"))
    overall.n = anova(mod.n, mod_null.n, test="LRT")

    #extract different p-values
    pval_overall.n = overall.n[2,5]
    pval_neg_vAtRisk.n = summary(mod_pairwise.n)$test$pvalues[1]
    pval_atRisk_vRA.n = summary(mod_pairwise.n)$test$pvalues[3]
    pval_neg_vRA.n = summary(mod_pairwise.n)$test$pvalues[2]

    toReport.n = c(pval_neg_vAtRisk.n, pval_atRisk_vRA.n, pval_neg_vRA.n, pval_overall.n) 
    names(toReport.n) = c("Negative: CCP+ vs CCP-", "Negative: RA vs CCP+", "Negative: RA vs CCP-", "Negative: overall p-value")
    
    toReport.all = c(toReport, toReport.n)
    return(toReport.all)
}

strat.results_pvals = t(apply(ACPAvars, 1, function(a) getStratifiedResults(a)))
rownames(strat.results_pvals) = ACPAvars[,1]

strat.results_pvals = as.data.frame(strat.results_pvals)
strat.results_pvals$Pos_FDR_CCPpos_vNeg = p.adjust(strat.results_pvals$`Positive: CCP+ vs CCP-`, method="BH")
strat.results_pvals$Neg_FDR_CCPpos_vNeg = p.adjust(strat.results_pvals$`Negative: CCP+ vs CCP-`, method="BH")

strat.results_pvals$Pos_FDR_RA_vCCPpos = p.adjust(strat.results_pvals$`Positive: RA vs CCP+`, method="BH")
strat.results_pvals$Neg_FDR_RA_vCCPpos = p.adjust(strat.results_pvals$`Negative: RA vs CCP+`, method="BH")

strat.results_pvals$Pos_FDR_RA_vNeg =  p.adjust(strat.results_pvals$`Positive: RA vs CCP-`, method="BH")
strat.results_pvals$Neg_FDR_RA_vNeg =  p.adjust(strat.results_pvals$`Negative: RA vs CCP-`, method="BH")

strat.results_pvals$Pos_RA_vCCPpos_pretty = paste0(formatC(strat.results_pvals$`Positive: RA vs CCP+`, format="e", digits=2), " (", round(strat.results_pvals$Pos_FDR_RA_vCCPpos, 3), ")")
strat.results_pvals$Neg_RA_vCCPpos_pretty = paste0(formatC(strat.results_pvals$`Negative: RA vs CCP+`, format="e", digits=2), " (", round(strat.results_pvals$Neg_FDR_RA_vCCPpos, 3), ")")

strat.results_pvals$Pos_RA_vCCPneg_pretty = paste0(formatC(strat.results_pvals$`Positive: RA vs CCP-`, format="e", digits=2), " (", round(strat.results_pvals$Pos_FDR_RA_vNeg, 3), ")")
strat.results_pvals$Neg_RA_vCCPneg_pretty = paste0(formatC(strat.results_pvals$`Negative: RA vs CCP-`, format="e", digits=2), " (", round(strat.results_pvals$Neg_FDR_RA_vNeg, 3), ")")

strat.results_pvals$Pos_CCPpos_vCCPneg_pretty = paste0(formatC(strat.results_pvals$`Positive: CCP+ vs CCP-`, format="e", digits=2), " (", round(strat.results_pvals$Pos_FDR_CCPpos_vNeg, 3), ")")
strat.results_pvals$Neg_CCPpos_vCCPneg_pretty = paste0(formatC(strat.results_pvals$`Negative: CCP+ vs CCP-`, format="e", digits=2), " (", round(strat.results_pvals$Neg_FDR_CCPpos_vNeg, 3), ")")

#rownames(strat.results_pvals[which(strat.results_pvals$`Negative: CCP+ vs CCP-`<0.05),])
```

Of the stratified results comparing CCP- and CCP+, nothing reached FDR < 0.05 criteria.  None of the positive 0401 genotype results even met a nominal p-value < 0.05 (minimum p-value for this strata is `r round(min(strat.results_pvals$"Positive: CCP+ vs CCP-"), 3)`).  However, for the negative 0401 genotype strata, there are **`r length(which(strat.results_pvals$"Negative: CCP+ vs CCP-"<0.05))`** ACPA peptides with a nominal p-value < 0.05.  They are: `r paste0(rownames(strat.results_pvals[which(strat.results_pvals$"Negative: CCP+ vs CCP-"<0.05),]), collapse=", ")`.

## 0401 Positive Results Table

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}
getNs = function(ACPA){
  tmp = forAnalysis_3SD
  colnames(tmp)[which(colnames(tmp)==ACPA)]="out"
  tmp.pos = tmp[which(tmp$geno0401_bi==1),]
  tmp.neg = tmp[which(tmp$geno0401_bi==0),]
  
  Ns.pos = table(tmp.pos$out, tmp.pos$Analyze)
  if(all(nrow(Ns.pos)==1 & rownames(Ns.pos)=="0")){want.pos = c("0 (0%)", "0 (0%)", "0 (0%)")}else{
    N_1.pos = Ns.pos[2,]
  percents.pos = paste0("(", round(100*Ns.pos/colSums(Ns.pos), 2)[2,], "%)")
  Ns_pos_per.pos = cbind(N_1.pos, percents.pos)
  want.pos = apply(Ns_pos_per.pos, 1, function(a) paste0(a[1], " ", a[2]))
  }
  
  Ns.neg = table(tmp.neg$out, tmp.neg$Analyze)    
  if(all(nrow(Ns.neg)==1 & rownames(Ns.neg)=="0")){want.neg = c("0 (0%)", "0 (0%)", "0 (0%)")}else{
  N_1.neg = Ns.neg[2,]
  percents.neg = paste0("(", round(100*Ns.neg/colSums(Ns.neg), 2)[2,], "%)")
  Ns_pos_per.neg = cbind(N_1.neg, percents.neg)
  want.neg = apply(Ns_pos_per.neg, 1, function(a) paste0(a[1], " ", a[2]))}

  names(want.pos) = paste0("0401 Positive:", c("CCP_Neg_Baseline", "CCP_Pos_Baseline", "RA_Baseline"))
  names(want.neg) = paste0("0401 Negative:", c("CCP_Neg_Baseline", "CCP_Pos_Baseline", "RA_Baseline"))

  want = c(want.pos, want.neg)
  return(want)
}
toPrint.strat = t(apply(as.matrix(ACPAvars), 1, function(a) getNs(a)))
rownames(toPrint.strat) = ACPAvars

toPrint.pos = toPrint.strat[,grep("Positive", colnames(toPrint.strat))]
toPrint.pos = merge(toPrint.pos, strat.results_pvals[,c(19,17)], by=0)
colnames(toPrint.pos) = c("ACPA Feature", "Anti-CCP3", "At-Risk", "Early RA", "At-Risk vs Anti-CCP3 p-value (FDR)", "Early RA vs At-Risk p-value (FDR)")

toPrint.neg = toPrint.strat[,grep("Negative", colnames(toPrint.strat))]
toPrint.neg = merge(toPrint.neg, strat.results_pvals[,c(20,18)], by=0)
colnames(toPrint.neg) = c("ACPA Feature", "Anti-CCP3", "At-Risk", "Early RA", "At-Risk vs Anti-CCP3 p-value (FDR)", "Early RA vs At-Risk p-value (FDR)")

datatable(toPrint.pos,rownames=FALSE)
```

## 0401 Negative Results Table

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}

datatable(toPrint.neg,rownames=FALSE)
```

## Heatmaps

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}
toPlot = forAnalysis_3SD[order(forAnalysis_3SD$Analyze),]
rownames(toPlot) = toPlot$LabId
toPlot.pos = toPlot[which(toPlot$geno0401_bi==1),]
toPlot.neg = toPlot[which(toPlot$geno0401_bi==0),]

toPlot.pos = toPlot.pos[,-c(1:3, 5:8)]
toPlot.neg = toPlot.neg[,-c(1:3, 5:8)]

toPlot.neg.neg = as.matrix(t(toPlot.neg[which(toPlot.neg$Analyze=="CCP_Neg_Baseline"),-1]))
toPlot.neg.pos =  as.matrix(t(toPlot.neg[which(toPlot.neg$Analyze=="CCP_Pos_Baseline"),-1]))
toPlot.neg.RA =  as.matrix(t(toPlot.neg[which(toPlot.neg$Analyze=="RA_Baseline"),-1]))


toPlot.pos.neg = as.matrix(t(toPlot.pos[which(toPlot.pos$Analyze=="CCP_Neg_Baseline"),-1]))
toPlot.pos.pos =  as.matrix(t(toPlot.pos[which(toPlot.pos$Analyze=="CCP_Pos_Baseline"),-1]))
toPlot.pos.RA =  as.matrix(t(toPlot.pos[which(toPlot.pos$Analyze=="RA_Baseline"),-1]))
```


**0401 Positive Strata**

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}
hp.pos = Heatmap(toPlot.pos.neg, col=c("darkblue", "red"), column_title="Anti-CCP3\n(n=33)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10)) + 
Heatmap(toPlot.pos.pos, col=c("darkblue", "red"), column_title="At-Risk\n(n=36)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10)) +
Heatmap(toPlot.pos.RA, col=c("darkblue", "red"), column_title="Early RA\n(n=24)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10))
hp.pos
#png(file="R:/TIP-RA/LaurenV/baselineCrossSection/manuscript/ACPA_heatmap.0401pos.png", res=600,width=480*12, height=480*7)
#hp.pos
#dev.off()
```

**0401 Negative Strata**

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}
hp.neg = Heatmap(toPlot.neg.neg, col=c("darkblue", "red"), column_title="Anti-CCP3\n(n=139)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10)) + 
Heatmap(toPlot.neg.pos, col=c("darkblue", "red"), column_title="At-Risk\n(n=71)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10)) +
Heatmap(toPlot.neg.RA, col=c("darkblue", "red"), column_title="Early RA\n(n=38)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10))
hp.neg
#png(file="R:/TIP-RA/LaurenV/baselineCrossSection/manuscript/ACPA_heatmap.0401neg.png", res=600,width=480*12, height=480*7)
#hp.neg
#dev.off()
```

# Alternative Heatmap

This is an alternative heatmap.  This has all samples, but added a bar on top of each to signify if the same is genotype 0401 positive (yellow) or 0401 negative (green).   

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.height = 4, fig.width = 6.5}
toPlot = forAnalysis_3SD[order(forAnalysis_3SD$Analyze),]
rownames(toPlot) = toPlot$LabId
toPlot = toPlot[,-c(1:2, 5:8)]
toPlot = toPlot[order(toPlot$Analyze, toPlot$geno0401_bi),]


toPlot.neg = as.matrix(t(toPlot[which(toPlot$Analyze=="CCP_Neg_Baseline"),-c(1:2)]))
toPlot.pos =  as.matrix(t(toPlot[which(toPlot$Analyze=="CCP_Pos_Baseline"),-c(1:2)]))
toPlot.RA =  as.matrix(t(toPlot[which(toPlot$Analyze=="RA_Baseline"),-c(1:2)]))

geno_col_fun = colorRamp2(c(0, 1), c("#61D04F" , "#F5C710")) 

ha.neg =  HeatmapAnnotation(geno0401=anno_simple(toPlot[which(toPlot$Analyze=="CCP_Neg_Baseline"),1], col = geno_col_fun), annotation_name_side = "left", annotation_name_gp = gpar(fontsize = 5), height = unit(2.5/25, "cm"))
ha.pos =  HeatmapAnnotation(geno0401=anno_simple(toPlot[which(toPlot$Analyze=="CCP_Pos_Baseline"),1], col = geno_col_fun), show_annotation_name = FALSE, height = unit(2.5/25, "cm"))
ha.RA =  HeatmapAnnotation(geno0401=anno_simple(toPlot[which(toPlot$Analyze=="RA_Baseline"),1], col = geno_col_fun), show_annotation_name = FALSE, height = unit(2.5/25, "cm"))

hp = Heatmap(toPlot.neg, col=c("darkblue", "red"), column_title="Anti-CCP3\n(n=172)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10), top_annotation = ha.neg) + 
Heatmap(toPlot.pos, col=c("darkblue", "red"), column_title="At-Risk\n(n=97)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10), top_annotation = ha.pos) +
Heatmap(toPlot.RA, col=c("darkblue", "red"), column_title="Early RA\n(n=62)",show_row_names = TRUE, cluster_columns = FALSE, cluster_rows=TRUE, show_row_dend = FALSE, row_names_max_width = unit(2.5, "cm"), show_column_names = FALSE, show_heatmap_legend = FALSE, row_names_gp = gpar(fontsize = 5), column_title_gp = gpar(fontsize = 10), top_annotation = ha.RA)
hp

#png(file="R:/TIP-RA/LaurenV/baselineCrossSection/manuscript/ACPA_heatmap_v2.png", res=600,width=480*12, height=480*7)
#hp
#dev.off()
```
